---
title: "ITEC 621 - Homework 1 - R and Stats Refresher"
subtitle: "Kogod School of Business"
author: "O'Malley, Conie"
date: "19 Jan 2025"
output:
  word_document:
   toc: true
   toc_depth: 2
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=T, warning=F, message=F)
library(tidyverse)
```

## Q1 - Functions and Loops (12.5 pts.)

Write a simple R function named `area()` that takes 2 values as input parameters (representing the two sides of a rectangle) and returns the product of the two values (representing the rectangle's area). 

Then use the functions `print()` and `paste()` to output this result: **"The area of a rectangle of sides 6x4 is 24"**, where 24 is calculated with the `area()` function you just created (not a fixed number). That is, the `area()` function you created, **must appear** inside the `print(paste())` function.

```{r area}
area <- function(x, y){ # create a function with 2 input parameters
  z = x*y # calculate area
  print(paste("The area of a rectangle of sides", x, "x", y, "is", z)) # print phrase with calculations
}

area(6,4) # function test
```

Write a simple **for loop** for i from **1 to 10** to compute **squared** values for the numbers 1 through 10. In each loop pass, compute the square of i (i.e., `i^2`) and in each case, display **exactly** "The square of 1 is 1" for i=1, "The square of 2 is 4" for i=2, and so on. You must use the functions `pring(paste())` and the formula `i^2` to display your results. 

```{r Loop}
for (i in 1:10){ # for loop range
  a = i**2 # calculate squares
  print(paste("The square of", i, "is", a)) # print phrase with calculations 
}
```

## Q2 - Data Frames and Vectors (12.5 pts.)

We suspect that income and loan balance may predict the credit rating of a bank's customer. Let's explore. Copy the Credit.csv data file to your working directory, then:

Read the Credit.csv data table into a data frame named "Credit" (tip: use the `read.table()` function with the appropriate `header=` and `sep=` attributes)

Display (only) the **first 10 rows** of (only) the Rating, Income and Balance columns. Use the `cbind()` function to bind the 3 columns and ensure that your columns have labels (e.g. "Rating"=Credit$Rating[1:10], etc.). 

```{r ReadCSV}
read.table("Credit.csv", header = TRUE, sep = ",") -> Credit # read data into table

cbind("Rating" = Credit$Rating[1:10], # display required columns
      "Income" = Credit$Income[1:10],
      "Balance" = Credit$Balance[1:10])
```

Then, display the object class for the Credit data frame and for the vectors Gender (i.e., `Credit$Gender`), Income and Cards

```{r Class}
class(Credit) # obtain class of Credit
class(Credit$Gender) # obtain class of Gender
class(Credit$Income) # obtain class of Income
class(Credit$Cards) # obtain class of Cards
```

Finally, create a vector named **Rating.vect** with data from the Rating column and display the first 6 values of this vector.

```{r Vector}
Rating.vect <- Credit$Rating # assign vector
Rating.vect[1:6] # display first 6 values
```

## Q3. Descriptive Analytics (12.5 pts.)

Let's analyze the data quantitatively. Compute the **mean, minimum, maximum, standard deviation** and **variance** for all the values in this Rating.vect vector. Store the results in variables named **rtg.mean, rtg.min, rtg.min, rtg.stdev** and **rtg.var** respectively. Then use the `c()` function to concatenate the 5 results into a vector, but name each of the values accordingly (e.g., "Mean Rtg"=rtg.mean). Also, enclose the `c()` function inside the `print()` function and use `digits=5` to limit the significant digits to display (to 2 decimals) (e.g., `print(c(...), digits=5)`)

```{r Descriptives}
rtg.mean <- mean(Rating.vect)
rtg.min <- min(Rating.vect)
rtg.max <- max(Rating.vect)
rtg.stdev <- sd(Rating.vect)
rtg.var <- var(Rating.vect)
print(c("Mean Rtg" = rtg.mean, "Min Rtg" = rtg.min, "Max Rtg" = rtg.max, "StDev Rtg" = rtg.stdev, "Var Rtg" = rtg.var), digits = 5)
```

Divide the plot output into 1 row and 2 columns to display 2 graphs side by side using `par(mfrow=c(1,2))`. Display a histogram for **Credit Rating**, with the main title **Credit Rating Histogram** and X label **Rating**. Since the histogram is a bit skewed to the right, display a histogram for the **Log of Credit Rating**, with the main title **Log of Credit Rating Histogram** and X label **Log Rating**. Tip: use the `log()` function (lower case) to log variables. Then, reset the graph layout to 1 by 1 using par(mfrow=c(1,1)).

**Technical Note:** Notice that we divide the plot window to display 2 graphs side by side and then reset to the normal window to a single plot, and we then do this repeatedly for every exercise below. This is only necessary so that when you press the play icon for the chunk, you can see the 2 graphs side by side. If you don't want to do this repeatedly, you could set the window for the first pair of plots and reset it after the last pair of plots. The play icon will not give you side by side graphs, but the knitted document should show the graphs side by side. Can you figure out why? (no need to answer, just think)

**Technical note:** also note in the HW template that I have sized the graphs to `fig.width=10`, `fig.height=6`. You can change this if you prefer a different graph size.

```{r Histogram, fig.width=10, fig.height=6}
#par(mfrow=c(1,2)) # set parameters
par(mfrow=c(1,1)) # reset parameters

hist(Credit$Rating, main = "Credit Rating Histogram", xlab = "Rating", ylab = NULL) # initial histogram

hist(log(Credit$Rating), main = "Log of Credit Rating Histogram", xlab = "Log Rating", ylab = NULL) # log histogram to address skewness
```

## Q4. Visual Analytics (12.5 pts.)

Let's do some visual inspection of the data. Divide the plot output into 1 row and 2 columns to display 2 graphs side by side using `par(mfrow=c(1,2))`. Then display a box plot for **Rating** by **Gender**, and then another box plot for **Balance** by **Student**. Then, reset the graph layout to 1 by 1 using par(mfrow=c(1,1)).

```{r Boxplots, fig.width=10, fig.height=6}
#par(mfrow=c(1,2)) # set parameters
par(mfrow=c(1,1)) # reset parameters

boxplot(Credit$Rating~Credit$Gender, main = "Box Plot by Gender", xlab = "Gender", ylab = "Credit Rating") # gender boxplot
boxplot(Credit$Rating~Credit$Student, main = "Box Plot by Student", xlab = "Student", ylab = "Credit Rating") # student boxplot
```

Divide again the plot output into 1 row and 2 columns to display 2 graphs side by side using `par(mfrow=c(1,2))`. Then plot Credit **Rating** (Y axis) against **Income** (X axis), with respective labels **Income** and **Credit Rating**. Then another plot for Credit **Rating** (Y axis) against **Balance** (X axis), with respective labels **Credit Rating** and **Balance**. Then, reset the graph layout to 1 by 1 using par(mfrow=c(1,1)). 

```{r ScatterPlot, fig.width=10, fig.height=6}
#par(mfrow=c(1,2)) # set parameters
par(mfrow=c(1,1)) # reset parameters
plot(Credit$Income, Credit$Rating, main = "Income vs. Credit Rating Plot", xlab = "Income", ylab = "Credit Rating") # income vs. credit rating plot
plot(Credit$Balance, Credit$Rating, main = "Balance vs. Credit Rating Plot", xlab = "Income", ylab = "Credit Rating") # balance vs. credit rating plot
```

Again, divide the plot output into 1 row and 2 columns to display 2 graphs side by side using `par(mfrow=c(1,2))`. Display a qqplot (using the `qqnorm()` function and `qqline()`) for the credit rating variable, with the main title **Credit Rating QQ PLot** (use the `main=` attribute). Since the QQ Plot shows some deviation from the normal distribution line, display a QQ Plot for the **Log of Credit Rating**, with the main title **Log of Credit Rating QQ PLot** and X label **Log Rating**. Then, reset the graph layout to 1 by 1 using `par(mfrow=c(1,1))`. It looks more normally distributed, right?

```{r Normality, fig.width=10, fig.height=6}
#par(mfrow=c(1,2)) # set parameters
par(mfrow=c(1,1)) # reset parameters

qqnorm(Credit$Rating, main = "Credit Rating QQ Plot", xlab = "Theoretical Quantiles", ylab = "Sample Quantiles")
qqline(Credit$Rating, col = "red")

qqnorm(log(Credit$Rating), main = "Log of Credit Rating QQ Plot", xlab = "Log Rating", ylab = "Sample Quantiles")
qqline(log(Credit$Rating), col = "red")
```

## Q5. {ggplot} Graphs (12.5 pts.)

**{ggplot2}** is one of the best visual analytic packages out there. Let's explore it briefly. Load the **{ggplot2}** library.

***Note:*** `ggplot()` renders one plot at a time, so there is no need to split the plot window into more than one row or column. 

Then, use the `ggplot()` function to draw a histogram of the **Rating** variable and then draw another histogram for **log(Rating)**. Tip: you need to use `ggplot(Credit)` first to point to the data set you want to use. You then need to add the necessary graph attributes with the **+** operator. In this case you need to use the `geom_histogram(aes(etc.))` graph attribute and aesthetic.

Then, use the `ggplot()` function again to draw a couple of scatter plots. In the first one, plot **Income** (x axis) against **Rating** (y axis), but separated by **Gender**. Tip, use `ggplot(Credit, aes(x=Income, y=Rating))` to use the 2 variable plot aesthetic with the **Credit** data set. Then add `geom_point()` to draw a scatter plot, and then add `facet_wrap(~Gender)` to draw two **facets by Gender**.

Then draw a similar scatter plot, but this time use **Balance** (x axis) by **Rating** (y axis), faceted by **Student**.

```{r ggplot, fig.width=10, fig.height=6}

```

## Q6. Linear Regression Models (12.5 pts.)

Fit a **small** linear regression model object with the `lm()` function to predict credit **Rating** using **Income** and **Balance** as predictors. Name the resulting linear model **fit.small**. Then display the `summary()` results. 

```{r fit.small}

```

Now fit a larger linear regression model, same as **fit.small**, but add 2 more predictors this time, **Age** and **Gender**. Name the resulting linear model **fit.large**. Then display the `summary()` results. 

```{r fit.large}

```

## Q7. ANOVA (12.5 pts.)

Do an `anova()` test to evaluate if **fit.large** has significantly more predictive power than **fit.small**. 

Then provide a brief **interpretation** of your results for Q5, Q6 and Q7. Which predictors are significant? which are not? and why? which of the two models is preferred? why?

```{r anova}

# IMPORTANT: I'm writing my commentary inside the R code chunk with a # text entry, only because I want to hide it for the homework. Once I turn the echo on with echo=T to show the code you will see my answer. However, you should NEVER write your interpretation reports inside the R code chunk. It looks unprofessional and some times {knitr} does not knit it well. Write all your answers outside of the code chunk.
```

## Q8. Regression Plots (12.5 pts.)

As you should know, the linear model object contains 4 plots. But in this exercise we will only plot the first 2. So, divide the plot window into 1 rows and 2 columns to display the 2 plots side by side. Then plot the **fit.large** linear model object, but use the attribute `which=1`. Then plot it again, but using `which=2` instead. Then reset the plot window to a single graph as you did earlier.

Then provide a brief interpretation of what you see in the two graphs and what issues the may entail.

```{r regPlots, fig.width=10, fig.height=6}

```

